Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-05 drafts/hybrid_zone.tex
YB edited through Lineages at the selected locus

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation. % too much `understanding' here
Here, we use a combination of analytical theory and explicit forward simulations
to show how selection against hybrid genotypes impacts patterns of introgression  % too much `understanding' here
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  physical scale of $1/T$, where $T$ is the age of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location. 
The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}
}

@@

As in \citet{sedghifar2015spatial}, 
we gain insight by considering where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to a spatial position $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same;
recombination events may cause us to follow more than one lineage back to determine the block's ancestry,
and the block will be of ancestry $A$ if all lineages trace back to the $A$ side.
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
As the identity of the selected allele determines how selection acts on the haplotype,
and linked variation can only move between the $A$ and the $B$ background in heterozygotes,
a key factor in these models is the density and fecundity of heterozygotes.

>>>>

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation. % too much `understanding' here
Here, we use a combination of analytical theory and explicit forward simulations
to show how selection against hybrid genotypes impacts patterns of introgression  % too much `understanding' here
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  physical scale of $1/T$, where $T$ is the age of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location. 
Our findings suggest that the density of ancestry junctions provide a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}
}

@@

As in \citet{sedghifar2015spatial}, consider where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to a spatial position $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same; 
	this occurs if there is no recombination in this block, or if all lineages generated by recombination events 
	in this block trace back to the $A$ side. 
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
Because the identity of the selected allele determines how selection acts on the haplotype,
and because linked variation can only move between the $A$ and the $B$ background in heterozygotes, 
a key factor in these models is the density and fecundity of heterozygotes.
